cardinalR: Generating interesting high-dimensional data structures

### 2025-09-23 paper-cardinalR.Rmd
Remove trefoil summary table

The Trefoil is a closed, nontrivial one-dimensional manifold embedded in $3\text{-}D$ or $4\text{-}D$ space (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:trefoil)"} else { "Figure \\@ref(fig:trefoil-proj)"})`). The trefoil features topological complexity in the form of self-overlaps, making it a valuable test case for evaluating the ability of non-linear dimension reduction methods to preserve global structure, loops, and embeddings in high-dimensional data. Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:trefoil-tb-html)', '\\@ref(tab:trefoil-tb-pdf)'))` summarizes these functions.

```{r}
trefoil_tb <- tibble(fun = c("gen_trefoil4d",
                            "gen_trefoil3d"), 
                        exp = c("Trefoil in $4\\text{-}D$.",
                                "Trefoil in $3\\text{-}D$."))
```

```{r trefoil-tb-html, eval=knitr::is_html_output()}
trefoil_tb |> 
  kable(caption = "cardinalR trefoil data generation functions", col.names = c("Function", "Explanation")) 
```

```{r trefoil-tb-pdf, eval=knitr::is_latex_output()}

trefoil_tb |> 
  kable(caption = "cardinalR trefoil data generation functions", format="latex", col.names = c("Function", "Explanation"), booktabs = T)  |>
  column_spec(1, width = "4cm") |>
  column_spec(2, width = "8cm")
```

>>>>

The Trefoil is a closed, nontrivial one-dimensional manifold embedded in $3\text{-}D$ or $4\text{-}D$ space (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:trefoil)"} else { "Figure \\@ref(fig:trefoil-proj)"})`). The trefoil features topological complexity in the form of self-overlaps, making it a valuable test case for evaluating the ability of non-linear dimension reduction methods to preserve global structure, loops, and embeddings in high-dimensional data.

### 2025-09-23 paper-cardinalR.tex
Updated abstract with all the comments
Remove trefoil summary table

\abstract{%
A high-dimensional dataset is one where each observation is described by many features, or dimensions, with associations between them. These datasets contain nonlinear manifolds in image and speech recognition, clusters in genomics and forensic analysis, and sparse distributions in text mining. We can generate data containing a variety of structures using mathematical functions and statistical distributions. High-dimensional data structures are useful for testing, validating, and improving algorithms used in dimensionality reduction, clustering, machine learning, and visualization. Their controlled complexity allows researchers to understand challenges posed in data analysis and helps to develop robust analytical methods across diverse scientific fields like bioinformatics, machine learning, and forensic science. Functions to generate a large variety of structures in high dimensions are organized into the R package \texttt{cardinalR}, along with some already generated examples, adding to the existing toolset of benchmark datasets for evaluating algorithms.
}

@@

\subsection{Installation}\label{installation}

The development version can be installed from GitHub:

\begin{verbatim}
pak::pak("JayaniLakshika/cardinalR")
\end{verbatim}

@@

The Trefoil is a closed, nontrivial one-dimensional manifold embedded in \(3\text{-}D\) or \(4\text{-}D\) space (Figure \ref{fig:trefoil-proj}). The trefoil features topological complexity in the form of self-overlaps, making it a valuable test case for evaluating the ability of non-linear dimension reduction methods to preserve global structure, loops, and embeddings in high-dimensional data. Table \ref{tab:trefoil-tb-pdf} summarizes these functions.

\caption{\label{tab:trefoil-tb-pdf}cardinalR trefoil data generation functions}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}>{\raggedright\arraybackslash}p{8cm}}
\toprule
Function & Explanation\\
\midrule
gen\_trefoil4d & Trefoil in \$4\textbackslash{}text\{-\}D\$.\\
gen\_trefoil3d & Trefoil in \$3\textbackslash{}text\{-\}D\$.\\
\bottomrule
\end{tabular}
\end{table}

>>>>

\abstract{%
A high-dimensional dataset is one where each observation is described by many features, or dimensions, with associations between them. These datasets contain nonlinear manifolds in image and speech recognition, clusters in genomics and forensic analysis, and sparse distributions in text mining. Data with a variety of structures can be generated using mathematical functions and statistical distributions to create test datasets. High-dimensional data structures are useful for testing, validating, and improving algorithms used in dimensionality reduction, clustering, machine learning, and visualization. Their controlled complexity allows researchers to understand challenges posed in data analysis and helps to develop robust analytical methods across diverse scientific fields like bioinformatics, machine learning, and forensic science. Functions to generate a large variety of structures in high dimensions are organized into the R package \texttt{cardinalR}, along with some already generated examples, adding to the existing toolset of benchmark datasets for evaluating algorithms.
}

@@

The \texttt{cardinalR} R package is available on GitHub at \href{https://github.com/JayaniLakshika/cardinalR}{JayaniLakshika/cardinalR}.

@@

The Trefoil is a closed, nontrivial one-dimensional manifold embedded in \(3\text{-}D\) or \(4\text{-}D\) space (Figure \ref{fig:trefoil-proj}). The trefoil features topological complexity in the form of self-overlaps, making it a valuable test case for evaluating the ability of non-linear dimension reduction methods to preserve global structure, loops, and embeddings in high-dimensional data.
